Effect of Beam Chromaticity on Foregrounds in Wide-Field Measurements of Redshifted 21~cm Power Spectra

### 2016-02-08 ForegroundSimulations/ms.tex

The High Frequency Structural Simulator (HFSS) was used to model the dish and its angular response used in this study. The HFSS model used prime focus optics with a 14~m faceted parabola with a spar f/D ratio of 0.32.  The model has a 1~m central hole in the aluminum surface which is filled with a dielectric material similar to dry soil. The feed used a full PAPER dipole inside of a cylindrical backplane, the backplane is modeled as an aluminum surface. For the metal parts of the dipole, the discs were modeled as aluminum at the actual size, and the arms and terminals were modeled as copper. Dielectric stand-offs and supporting members were included. For the calculations, one pair of arms was excited using a modal port. These models cover a frequency range of 100--200~MHz in intervals of 1~MHz. Please refer to DeBoer~et~al.~(2016) for a complete description of the dish model.

@@

We investigate the effects of spectral chromaticity of the power pattern in more detail below.

@@

Also shown are the tracks (black dots) of point sources brighter than 50~Jy in the LST range 0--12~hrs. The solid circle near the bottom denotes the south celestial pole. It is noted that there are a few bright point sources that spend a majority of time in this LST range near the northern and southern horizons which will contaminate the measured visibilities on north-south antenna spacings. Future simulations of the dish power pattern must take the bright foreground locations, especially near the horizon, into account.

@@

We simulate visibilities from our foreground model using three different models for the antenna power pattern -- {\it Airy} pattern, simulated dish, and a pattern at 150~MHz of the simulated dish assumed to be achromatic over the entire band.

Fig.~\ref{fig:asm-dps-beam-chromaticity-baselines} shows the delay power spectra of foregrounds obtained with the aforementioned models for power pattern. In all these panels, power spectra obtained with achromatic, {\it Airy} disk and simulated disk are shown in black, red, and blue respectively. The vertical gray lines denote the horizon limits for a 14.6~m antenna spacing. The full band delay power spectra of the foregrounds measured on a 14.6~m antenna spacing oriented eastward is shown in the top left panel. A clear trend of broadening of spillover-wings outside the horizon limits is seen with increasing chromaticity. For instance, the spillover from foreground delay power spectrum obtained with simulated dish pattern extends out to $k_\parallel \approx \pm 0.14\,h$~Mpc$^{-1}$, with the {\it Airy} pattern it extends out to $k_\parallel \approx \pm 0.08\,h$~Mpc$^{-1}$, while the achromatic beam extends it only out to $k_\parallel \approx \pm 0.05\,h$~Mpc$^{-1}$. Thus an increase in chromaticity extends the foreground spillover into higher $k_\parallel$-modes making them inaccessible for EoR H{\sc i} signal detection.

Following the foreground removal strategy, the top middle panel shows these foreground delay spectra deconvolved using the complex delay CLEAN algorithm \citep{tay99,par09,par12b}. The CLEAN window covers delay bins between the horizon delay limits and extends further by three delays bins on each side of the horizon limit where a delay bin is of size inverse of the bandwidth. The panel at the top right shows the residuals left behind after deconvolution.

% \begin{figure*}[htb]
% \centering
% \includegraphics[width=\linewidth]{asm_beam_chromaticity_bl2_LST_3.96_hrs_subband_150.0_MHz.eps}
% \caption{Effect of chromaticity of antenna power pattern on foreground delay power spectra for a 14.6~m baseline.}
% \label{fig:asm-dps-beam-chromaticity-14.6m-baseline}
% \end{figure*}

% \begin{figure*}[htb]
% \centering
% \includegraphics[width=\linewidth]{asm_beam_chromaticity_bl8_LST_3.96_hrs_subband_150.0_MHz.eps}
% \caption{Effect of chromaticity of antenna power pattern on foreground delay power spectra for a 29.2~m baseline.}
% \label{fig:asm-dps-beam-chromaticity-29.2m-baseline}
% \end{figure*}

@@

@@

@@

Owing to the chromaticity of power patterns, the spillover from foregrounds extends beyond the horizon limits, most severely in the simulated disk pattern. As a result the residuals have significant power remaining on either side of the horizon limits since the window for deconvolution is narrower than the extent of spillover along $k_\parallel$. The two peaks on either side is most severe for the simulated disk pattern followed by that for the {\it Airy} disk pattern and least for an achromatic power pattern.

We investigate the effect of chromaticity of power pattern in a foreground avoidance strategy that avoids the aforementioned deconvolution of delay spectrum. Application of window functions with high sidelobe suppression have been discussed for mitigating foreground contamination \citep{thy13}. In line with the foreground avoidance strategy, the visibilities from the full band foreground simulation (whose delay power spectrum is shown on the top left) are multiplied by a Blackman-Harris window function of 10~MHz effective bandwidth. The delay power spectra so produced are shown in the panel on the bottom left. Due to the decreased effective bandwidth the resolution of the delay power spectrum appears decreased. Hence, only the $|k_\parallel| \gtrsim 0.14\,h$~Mpc$^{-1}$ modes remain accessible. Due to the overall lower amplitude of the {\it Airy} power pattern relative to the simulated dish power pattern and the achromatic pattern, the sidelobe levels of foreground delay power spectra from an {\it Airy} pattern are lower than the other models by almost two orders of magnitude.

Following a delay-deconvolution based foreground removal strategy, we apply the Blackman-Harris window of effective bandwidth 10~MHz on the unsubtracted foreground residuals (top right panel) in frequency domain and obtain the delay power spectrum shown in the bottom right panel. It is noted that lesser the chromaticity the more effective the deconvolution process is in lowering the residuals in and around the horizon limits. For instance, the peak of the delay power spectrum using the achromatic beam is lowered by four orders of magnitude relative to the original peak whereas the residuals of delay power spectra with chromatic {\it Airy} and simulated power patterns is lowered only by two orders of magnitude. However, owing to the ineffective subtraction of the response extended beyond the horizon limits, the sidelobe levels have not been lowered significantly and are within an order of magnitude of each other.

In both strategies, the EoR H{\sc i} signal power spectrum obtained by simulations with 21cmFAST (Mesinger et al. 2008) is shown for reference (dashed black lines) in the bottom panels.

>>>>

The High Frequency Structural Simulator (HFSS) was used to model the dish and its angular response used in this study. The HFSS model used prime focus optics with a 14~m faceted parabola with a spar f/D ratio of 0.32.  The model has a 1~m central hole in the aluminum surface which is filled with a dielectric material similar to dry soil. The feed used a full PAPER dipole inside of a cylindrical backplane, the backplane is modeled as an aluminum surface. For the metal parts of the dipole, the discs were modeled as aluminum at the actual size, and the arms and terminals were modeled as copper. Dielectric stand-offs and supporting members were included. For the calculations, one pair of arms was excited using a modal port. These models cover a frequency range of 100--200~MHz in intervals of 1~MHz. Refer to DeBoer~et~al.~(2016) for a complete description of the dish model.

@@

@@

Also shown are the tracks (black dots) of point sources brighter than 50~Jy in the LST range 0--12~hrs. The solid circle near the bottom denotes the south celestial pole. It is noted that there are a few bright point sources that spend a majority of time in this LST range near the northern and southern horizons which will contaminate the measured visibilities on north-south antenna spacings. Simulations of the antenna power pattern must take the bright foreground locations, especially near the horizon, into account.

@@

In HERA-19 array layout configuration, there are 30 unique baseline vectors and 8 unique baseline lengths. Fig.~\ref{fig:asm-dps-beam-chromaticity-baselines} shows the delay power spectra of foregrounds on the 8 unique baseline lengths obtained with the aforementioned models for power pattern. In all these panels, the full-band foreground delay power spectra obtained with achromatic, {\it Airy} and chromatic beam patterns are shown in black, red, and blue respectively. The vertical dotted lines denote the horizon limits for baseline lengths indicated in each panel. A clear trend of broadening of spillover-wings outside the horizon limits is seen with increasing chromaticity. For instance, the spillover from foreground delay power spectrum obtained with chromatic beam pattern is restricted to $|k_\parallel| \lesssim 0.2\,h$~Mpc$^{-1}$, with the {\it Airy} pattern it is restricted to $|k_\parallel| \lesssim 0.15\,h$~Mpc$^{-1}$, while with the achromatic beam it is restricted to $|k_\parallel| \lesssim 0.12\,h$~Mpc$^{-1}$ even on longest baseline lengths. Thus an increase in chromaticity extends the foreground spillover into higher $k_\parallel$-modes making them inaccessible for EoR H{\sc i} signal detection.

Hence it is clearly demonstrated that higher the beam chromaticity, the farther the foreground contamination inherently extends beyond the horizon limits of the {\it foreground wedge}. The {\it Airy} pattern typically creates one secondary peak whereas the chromatic beam is found to produce multiple secondary peaks outside the {\it foreground wedge} on each side.

@@

It must be emphasized that such a significant spillover is not limited by the horizon limits of the {\it foreground wedge} because this is caused by spectral structure in the antenna beam pattern in addition to that caused by position-dependent geometric phase in the visibility spectra. As a result delay-based complex deconvolution techniques such as CLEAN \citep{tay99,par09,par12b} that rely on smoothness of foreground spectra and only spectral window shape will not have adequate information to accurately deconvolve intrinsic supra-horizon spillover arising from the chromaticity of the antenna beam. Thus, we leave delay power spectrum estimation using {\it foreground removal} strategy to future work and use {\it foreground avoidance} approach in this study.

We investigate the effect of chromaticity of power pattern in a {\it foreground avoidance} strategy that avoids the aforementioned deconvolution of delay spectrum. Application of window functions with high sidelobe suppression have been discussed for mitigating foreground contamination \citep{thy13}. In line with the {\it foreground avoidance} strategy, Fig.~\ref{fig:asm-dps-beam-chromaticity-baselines-150-subband} shows delay spectra after applying a {\it Blackman-Harris} window function extending 10~MHz on each side centered on 150~MHz with effective bandwidth of 10~MHz. The 8 panels correspond to those shown in Fig.~\ref{fig:asm-dps-beam-chromaticity-baselines}. As a result of narrowed subband, the resolution in delay space (and $k_\parallel$) has coarsened. The cyan and gray lines denote power spectra of two EoR models, namely, those obtained from 21cmFAST \citep{mes11} and \citet{lid08} respectively. These models will be denoted hereafter as AM and AL respectively.

@@

Firstly, it is noted that the coarsening of delay resolution significantly absorbs the distinct differences caused by beams of different chromaticities as seen in the full band foreground delay spectra outside the {\it foreground wedge} in Fig.~\ref{fig:asm-dps-beam-chromaticity-baselines}. Secondly, it is also found that barring the shortest baselines ($|\boldsymbol{b}| \leq $~29.2~m), which have the most severe foreground contamination especially from diffuse emission, foreground contamination on baselines longer than $\gtrsim 29.2$~m is lesser than EoR signal power for $|k_\parallel| \gtrsim 0.22\,h$~Mpc$^{-1}$ in either of the models.

Fig.~\ref{fig:asm-dps-beam-chromaticity-baselines-170-subband} is similar to Fig.~\ref{fig:asm-dps-beam-chromaticity-baselines-150-subband} except it is obtained with a subband {\it Blackman-Harris} window function centered on 170~MHz. In this subband relative to that centered on 150~MHz, the ratio of powers of the EoR signal and foregrounds is even higher outside the wedge thus indicating a stronger potential for direct detectability without the use of sophisticated power spectrum estimation techniques on almost all baseline lengths.

@@

It must be emphasized that these are obtained only with windowing techniques and do not represent the best sensitivity achievable with sophisticated tools such as optimal covariance-based weighting schemes \citep{liu14a,liu14b,ali15,dil15}. Thus assuming the data is limited by foregrounds and not by thermal noise, this demonstrates that even with no sophistication in power spectrum estimation, a direct detection based on simple windowing and delay spectrum technique is possible on a majority of HERA baselines and will further allow distinguishing between different EoR models. This finding holds even in the most conservative scenario where the modeled beam pattern with most severe chromaticity inherently results in significant and inherent foreground contamination outside the {\it foreground wedge}.

% Owing to the chromaticity of power patterns, the spillover from foregrounds extends beyond the horizon limits, most severely in the simulated disk pattern. As a result the residuals have significant power remaining on either side of the horizon limits since the window for deconvolution is narrower than the extent of spillover along $k_\parallel$. The two peaks on either side is most severe for the simulated disk pattern followed by that for the {\it Airy} disk pattern and least for an achromatic power pattern.

% We investigate the effect of chromaticity of power pattern in a foreground avoidance strategy that avoids the aforementioned deconvolution of delay spectrum. Application of window functions with high sidelobe suppression have been discussed for mitigating foreground contamination \citep{thy13}. In line with the foreground avoidance strategy, the visibilities from the full band foreground simulation (whose delay power spectrum is shown on the top left) are multiplied by a Blackman-Harris window function of 10~MHz effective bandwidth. The delay power spectra so produced are shown in the panel on the bottom left. Due to the decreased effective bandwidth the resolution of the delay power spectrum appears decreased. Hence, only the $|k_\parallel| \gtrsim 0.14\,h$~Mpc$^{-1}$ modes remain accessible. Due to the overall lower amplitude of the {\it Airy} power pattern relative to the simulated dish power pattern and the achromatic pattern, the sidelobe levels of foreground delay power spectra from an {\it Airy} pattern are lower than the other models by almost two orders of magnitude.

% Following a delay-deconvolution based foreground removal strategy, we apply the Blackman-Harris window of effective bandwidth 10~MHz on the unsubtracted foreground residuals (top right panel) in frequency domain and obtain the delay power spectrum shown in the bottom right panel. It is noted that lesser the chromaticity the more effective the deconvolution process is in lowering the residuals in and around the horizon limits. For instance, the peak of the delay power spectrum using the achromatic beam is lowered by four orders of magnitude relative to the original peak whereas the residuals of delay power spectra with chromatic {\it Airy} and simulated power patterns is lowered only by two orders of magnitude. However, owing to the ineffective subtraction of the response extended beyond the horizon limits, the sidelobe levels have not been lowered significantly and are within an order of magnitude of each other.

% In both strategies, the EoR H{\sc i} signal power spectrum obtained by simulations with 21cmFAST (Mesinger et al. 2008) is shown for reference (dashed black lines) in the bottom panels.
